What can we learn about life in the city by mining the Melbourne’s Open Data Platform?

### 2017-05-06 thesis/proposal_draft.Rmd
updated

---
title: "What can we learn about life in the city by mining the Melbourne’s Open Data Platform?"
subtitle: "Thesis Proposal"
date: April 2017
author: 
- "Gavin Chin"
- "supervised by Dianne Cook"
output:
  pdf_document: default
  html_document: default
---

@@

---------------------------------------------------------------------------------------

@@

```{r, echo = F, fig.align= "center", fig.width=3.5, fig.height=3.5, cache=TRUE}
ggmap(melb) + geom_point(aes(x = Longitude, y = Latitude), data = ped_loc, colour = "red", alpha = 0.7) +
              labs(title = "Melbourne Pedestrian Sensor Locations") +
              xlab(NULL) + ylab(NULL) +
              theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(hjust = 0.5),
                    axis.text.x = element_blank(), axis.text.y = element_blank(),
                    axis.ticks = element_blank())
```
\begin{center}
Figure 1:  Locations of all pedestrian sensors around Melbourne's CBD as at April 2017 - Plotted using `ggmap`, the red points map the locations of the sensors. The majority of the pedestrian sensors are placed in the eastern side of the CBD, particularly along Swanston St
\end{center}

@@

\begin{center}
Figure 2: Smoothed pedestrian counts for Melbourne Central by day of the week and month of the year
\end{center}

@@

\begin{center}
Figure 3: Pedestrian Counting System Visualisation from www.pedestrian.melbourne.vic.gov.au - The UI allows the user to choose specific dates and hours of the day to visualise on a 3D perspective map

@@

\begin{center}
Figure 4: An interactive data visualisation map created using Tableau by Boreak Silk. Average daily volumes are mapped to size, and colours map a positive or negative change in the average for the period compared to a specified reference period. Note the lack of a legend to help users understand how to interpret different aesthetic mapping of the visualisation.
\end{center}
_Source:https://public.tableau.com/profile/boreak.silk#!/vizhome/PedestrianactivityinMelbourne/Dailyvolumesandcomparison _

@@

The pedestrian data is presented as a time series. However, we face a major obstacle with pedestrian data: multiple seasonal patterns. Looking at the data on a month to month basis at a single sensor location, we have two main cycles: a daily sub-cycle within a weekly cycle.
```{r, message= FALSE, warning= FALSE, fig.height=2, echo = FALSE, fig.align= "center"}
bsplot <- ped_df %>% filter(Year == "2016", Month == "June", Mdate < 25, Sensor_ID == "13")
ggplot(bsplot) + geom_line(aes(x = MTime, y = Hourly_Counts)) +
      xlab("Hour of the Month") + ylab("Hourly Counts") + theme_light()
```
\begin{center}
Figure 5:  Sample of hourly counts of pedestrians at Flagstaff Station (June 2016)
\end{center}  
From _Figure 5_, we can observe multiple characteristics of the time series. Firstly, we see the weekly cycle, where weekdays (Monday to Friday) have lots of activity and weekends (Saturday and Sunday) are relatively quiet. This would be highly suggestive that Flagstaff Station is primarily used by city office workers. Another phenomenon we observe is the effect of a public holiday. Looking near the middle of the month (Monday, 13 June 2016), we have the Queen's Birthday public holiday. We must also note that, prior to the start of 2016, Flagstaff Station was not operating on weekends, nor public holidays.
```{r, message= FALSE, warning= FALSE, fig.height=3, echo = FALSE, fig.align= "center"}
ggplot(filter(ped_df, Sensor_ID == "13")) + geom_path(aes(x = as.integer(Time), y = Hourly_Counts, colour = Day), alpha = 0.2) +
      xlab("Hour of the Day") + ylab("Hourly Counts") + theme_light()

```
\begin{center}
 **Figure 6**:  Hourly counts at Flagstaff Station, with colour mapped to day of the week - There are two distinctly different daily patterns. The typical weekday traffic pattern has two peaks at 9am and 5pm, while it remains relatively quiet over the weekend as seen by the flatter lines. Also note that the second peak is lower than the first peak on average.
\end{center}

@@

The use of weather data to improve forecasting performance would be ideal, due to the availability of weather forecasts compared to many other forms of data. Machine learning techniques will also be explored. In particular, clustering of sensor locations may be useful in identifying patterns at certain types of location, such as tourist attractions and shopping districts.  Different models for different types of locations may be used, and clustering may give insights as to which type of model to use. For example, the major city loop train stations, Flinders St, Southern Cross and Melbourne Central may have similar patterns. Then, we would expect to be able to use the same parameterisation of the model to estimate. After a model has been established, we need to evaluate the performance of the model. For this, we have decided to minimise the RMSFE (root mean square forecast error):
$$ \text{RMSFE} = \sqrt{\frac{\sum_{t=1}^{T}(\hat{y_t} - y_t)^2}{T}}$$

@@

This proposal is written in R Markdown (`.Rmd`) using RStudio and the code used to produce the results outputted are embedded in this document. The packages used for this proposal are: `readr, dplyr, ggplot2, ggmap, knitr`
## References  
  1. City of Melbourne, _'About Melbourne Data'_, https://data.melbourne.vic.gov.au/about,  
   $\quad$ _accessed 18 April 2017_

@@

### Plots of time series at each sensor location over entire time span from 2009:  
```{r, echo = FALSE, cache=TRUE, out.height = "70%"}
ped_plot <- aggregate(x = ped_df$Hourly_Counts, by = list("Sensor_ID" = ped_df$Sensor_ID, "Year" = ped_df$Year, "Month" = ped_df$Month), FUN = sum)

@@

Looking at the plots of the monthly aggregates, we can inspect the data quickly for obvious outliers. After detrending the time series for both the deterministic trends (non-zero mean) and stochastic trends (weekly and daily cycles), we expect to have a stationary series. Anything outside these "expected" trends from the cycles could indicate an unobserved variable (such as weather) having a casual effect on the pedestrian count. Large "dips" in the series, or odd values over a long time span may be indicative of missing values which need to be imputed.

>>>>

---
title: "What can we learn about life in the city by mining the Melbourne’s Open Data Platform?"
subtitle: "Thesis Proposal"
date: April 2017
author: 
- "Gavin Chin"
- "supervised by Dianne Cook"
output:
  pdf_document: 
    keep_tex: yes
  html_document: default
---

@@

@@

```{r, echo = F, fig.align= "center", fig.width=3.5, fig.height=3.5, cache=TRUE}
ggmap(melb) + geom_point(aes(x = Longitude, y = Latitude), data = ped_loc, colour = "red", alpha = 0.7) +
              labs(title = "Melbourne Pedestrian Sensor Locations") +
              xlab(NULL) + ylab(NULL) +
              theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(hjust = 0.5),
                    axis.text.x = element_blank(), axis.text.y = element_blank(),
                    axis.ticks = element_blank())
```
\begin{center}
\textbf{Figure 1}:  Locations of all pedestrian sensors around Melbourne's CBD as at April 2017 - Plotted using `ggmap`, the red points map the locations of the sensors. The majority of the pedestrian sensors are placed in the eastern side of the CBD, particularly along Swanston St
\end{center}

@@

\begin{center}
\textbf{Figure 2}: Smoothed pedestrian counts for Melbourne Central by day of the week and month of the year
\end{center}

@@

\begin{center}
\textbf{Figure 3}: Pedestrian Counting System Visualisation from www.pedestrian.melbourne.vic.gov.au - The UI allows the user to choose specific dates and hours of the day to visualise on a 3D perspective map

@@

\begin{center}
\textbf{Figure 4}: An interactive data visualisation map created using Tableau by Boreak Silk. Average daily volumes are mapped to size, and colours map a positive or negative change in the average for the period compared to a specified reference period. Note the lack of a legend to help users understand how to interpret different aesthetic mapping of the visualisation.
\end{center}
_Source:https://public.tableau.com/profile/boreak.silk#!/vizhome/PedestrianactivityinMelbourne/Dailyvolumesandcomparison _

@@

The pedestrian data is presented as a time series. However, we face a major obstacle with pedestrian data: multiple seasonal patterns. Looking at the data on a month to month basis at a single sensor location, we have two main cycles: a daily sub-cycle within a weekly cycle.
```{r, message= FALSE, warning= FALSE, fig.height=2, echo = FALSE, fig.align= "center"}
bsplot <- ped_df %>% filter(Year == "2016", Month == "June", Mdate < 25, Sensor_ID == "13")
ggplot(bsplot) + geom_line(aes(x = MTime, y = Hourly_Counts)) +
      xlab("Hour of the Month") + ylab("Hourly Counts") + theme_light()
```
\begin{center}
\textbf{Figure 5}:  Sample of hourly counts of pedestrians at Flagstaff Station (June 2016)
\end{center}  
From _Figure 5_, we can observe multiple characteristics of the time series. Firstly, we see the weekly cycle, where weekdays (Monday to Friday) have lots of activity and weekends (Saturday and Sunday) are relatively quiet. This would be highly suggestive that Flagstaff Station is primarily used by city office workers. Another phenomenon we observe is the effect of a public holiday. Looking near the middle of the month (Monday, 13 June 2016), we have the Queen's Birthday public holiday. We must also note that, prior to the start of 2016, Flagstaff Station was not operating on weekends, nor public holidays.
```{r, message= FALSE, warning= FALSE, fig.height=3, echo = FALSE, fig.align= "center"}
ggplot(filter(ped_df, Sensor_ID == "13")) + geom_path(aes(x = as.integer(Time), y = Hourly_Counts, colour = Day), alpha = 0.2) +
      xlab("Hour of the Day") + ylab("Hourly Counts") + theme_light()

```
\begin{center}
 \textbf{Figure 6}:  Hourly counts at Flagstaff Station, with colour mapped to day of the week - There are two distinctly different daily patterns. The typical weekday traffic pattern has two peaks at 9am and 5pm, while it remains relatively quiet over the weekend as seen by the flatter lines. Also note that the second peak is lower than the first peak on average.
\end{center}

@@

The use of weather data to improve forecasting performance would be ideal, due to the availability of weather forecasts compared to many other forms of data. Machine learning techniques will also be explored. In particular, clustering of sensor locations may be useful in identifying patterns at certain types of location, such as tourist attractions and shopping districts.  Different models for different types of locations may be used, and clustering may give insights as to which type of model to use. For example, the major city loop train stations, Flinders St, Southern Cross and Melbourne Central may have similar patterns. Then, we would expect to be able to use the same parameterisation of the model to estimate. After a model has been established, we need to evaluate the performance of the model. For this, we have decided to minimise the RMSFE (root mean square forecast error). Assuming that we estimate the model with $T$ observations, and we have $k$ points to forecast (so we need to have $T+k$ observations available), then our RMSFE is:
$$ \text{RMSFE} = \sqrt{\frac{\sum_{t=T+1}^{T+k}(\hat{y_t} - y_t)^2}{k}}$$

@@

This proposal is written in R Markdown (`.Rmd`) using RStudio and the code used to produce the results outputted are embedded in this document. The packages used for this proposal are: `readr, dplyr, ggplot2, ggmap, knitr`

## References  
  1. City of Melbourne, _'About Melbourne Data'_, https://data.melbourne.vic.gov.au/about,  
   $\quad$ _accessed 18 April 2017_

@@

### Plots of time series at each sensor location over entire time span from 2009:

```{r, echo = FALSE, cache = TRUE, fig.height = 2.5}
ped_plot <- aggregate(x = ped_df$Hourly_Counts, by = list("Sensor_ID" = ped_df$Sensor_ID, "Year" = ped_df$Year, "Month" = ped_df$Month), FUN = sum)

@@

```
Looking at the plots of the monthly aggregates, we can inspect the data quickly for obvious outliers. After detrending the time series for both the deterministic trends (non-zero mean) and stochastic trends (weekly and daily cycles), we expect to have a stationary series. Anything outside these "expected" trends from the cycles could indicate an unobserved variable (such as weather) having a casual effect on the pedestrian count. Large "dips" in the series, or odd values over a long time span may be indicative of missing values which need to be imputed.
